A Transactional Memory System for Parallel Python

### 2014-06-06 talk/dls2014/paper/paper.tex
add abstract

\begin{abstract}
\remi{missing abstract}
\end{abstract}

@@

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is
therefore beneficial also for environments with a small number of
threads or cores. It is also used to implement atomic blocks into the
Python language to provide a backwards compatible, composable
synchronisation mechanism for threads.

@@

To obtain an overall performance that is competitive with the
best-performing Python systems (Jython, PyPy), integration of the
STM-based approach with a JIT compiler is necessary. Once this
integration matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards
parallelizing Python.

>>>>

\begin{abstract}

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language implementations is the use of
a single global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, they can not make use of the increasing parallelism in
current hardware.

In this paper, we present a software transactional memory (STM) system
specifically designed for the purpose of replacing the GIL in dynamic
language interpreters. The key idea is a close integration of STM with
the garbage collector to provide atomicity, synchronisation, and
memory management in a unified way.  As a user abstraction for a
scalable synchronisation mechanism, we implement atomic blocks by
exposing part of the STM system to applications.  The described STM
uses a combination of features present in current CPUs to lower its
overhead so that it can beat the performance of a GIL supported Python
interpreter already on 2 threads.

\end{abstract}

@@

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is
therefore beneficial also for environments with a small number of
threads or cores. It is also used to implement atomic blocks for the
Python language to provide a backwards compatible, composable
synchronisation mechanism for threads.

@@

To obtain an overall performance that is competitive with the
best-performing Python systems (Jython, PyPy), integration of the
STM-based approach with a JIT compiler is necessary. Once this
integration matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards
parallelising Python.
